---
title: "Appendix 2: Using cobalt with Clustered, Multiply Imputed, and Other Segmented Data"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette

vignette: >
  %\VignetteIndexEntry{Appendix 2: Using cobalt with Clustered, Multiply Imputed, and Other Segmented Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
  knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 5)
#knitr::opts_chunk$set(eval = FALSE)
```

This is a guide for the use of `cobalt` with more complicated data than is typical in studies using propensity scores and similar methods. In particular, this guide will explain `cobalt`'s features for handling multilevel or grouped data and data arising from multiple imputation. The features described here set `cobalt` apart from other packages that assess balance because they exist only in `cobalt`. It will be assumed that the basic functions of `cobalt` are understood; this guide will only address issues that are unique to these data scenarios.

## `cobalt` and Segmented Data

First, let's understand segmented data. Segmented data arises when the data involved in balance assessment needs to be split into segments to appropriately assess balance. These scenarios include clustered (e.g., multilevel) data, in which case balance should be assessed within each cluster; data arising from a sequential study, in which case balance should be assessed at each time point; multi-category treatments, in which case balance should be assessed for each pair of treatments; and multiply imputed data, in which case balance should be assessed within each imputation. `cobalt` can handle all these scenarios simultaneously, but how it does so may be a little complicated. This vignette explains how these scenarios are handled.

At the core is the idea that the most basic unit of balance assessment is a balance statistic for a covariate For binary treatments or pairs of treatment levels, this can be the (standardized) mean difference, variance ratio, or Kolmogorov-Smirnoff (KS) statistic. For continuous treatments, this is the treatment-covariate correlation. These statistics are generated by `bal.tab()` and can be plotted using `love.plot()` when the data are not segmented. When the data are segmented, these statistics need to be generated within each segment. When the segmentation occurs in several ways in the same dataset (e.g., with clustered and multiply imputed data, or with longitudinal data with multi-category treatments), balance assessment should reflect each layer of segmentation.

Although the idea of simply splitting data into segments is simple, there are a few options and limitations in `cobalt` that are important to consider. The basic idea is the same regardless of how the data are segmented: for each layer of segmentation, balance is assessed within segments of that layer, and the layers stack heirarchically. For example, for clustered and multiply imputed data, first the data are split by cluster; within each cluster, the data are split by imputation; balance statistics are computed within each imputation within each cluster. In some cases, a summary of balance across segments can be produced to simplify balance assessment. Matching and weighting are compatible with segmented data, but subclassification is its own special form of segmentation that is treated differently and will not be considered here. 

Each of `cobalt`'s primary functions (`bal.tab()`, `bal.plot()`, and `love.plot()`) have features to handle segmented data sets. The following sections describe for each data scenario the relevant features of each function. We'll take a look at a few common examples of segmented data: clustered data, multiply imputed data, and multi-category and multiply imputed data.

## Clustered Data

In clustered data, the data set must contain a variable denoting the group each individual belongs to. This may be a group considered a nuisance that must be accounted for to eliminate confounding (e.g., hospitals in a multi-site medical treatment study), or a group of concern for effect moderation (e.g., race or gender). In the examples below, we will imagine that we are interested in the ATT of `treat` on `re78` stratified by `race`. Thus, we will condition on the propensity score within each cluster.

First, let's estimate propensity scores and perform matching within each race group. We can do this by performing separate analyses within each cluster, but we can also use exact matching in `MatchIt` to ensure matches occur within clusters. It is important to note that this analysis does not necessarily represent a sound statistical analysis and is being used for illustrative purposes only.

```{r}
library("MatchIt"); library("cobalt")
data("lalonde", package = "cobalt")

m.out <- matchit(treat ~ race*(age + educ + married + nodegree + re74 + re75), 
                 data = lalonde, method = "nearest", exact = "race", 
                 replace = TRUE, ratio = 2)
```

### `bal.tab()`

The output produced by `bal.tab()` with clustered data contains balance tables for each cluster and a summary of balance across clusters. To use `bal.tab()` with groups, there are four arguments that should be considered. These are `cluster`, `which.cluster`, `cluster.summary`, and `cluster.fun`. 

* `cluster` is a vector of group membership for each unit or the name of a variable in a provided data set containing group membership.

* `which.cluster` determines for which clusters balance tables are to be displayed, if any. (Default: display all clusters)

* `cluster.summary` determines whether the cluster summary is to be displayed or not.
The arguments are in addition to the other arguments that are used with `bal.tab()` to display balance. (Default: display the cluster summary)

* `cluster.fun` determines which function(s) are used to combine balance statistics across clusters for the cluster summary. (Default: when `abs = FALSE`, minimum, mean, and maximum; when `abs = TRUE`, mean and maximum)

Lets examine balance on our data within each race group.

```{r}
bal.tab(m.out, cluster = "race")
```

First we see balance tables for each cluster. These are the same output we would see if we use `bal.tab()` for each cluster separately (e.g., using the `subset` argument). All the commands that work for `bal.tab()` also work here with the same results, except that balance tallies and the variable with the greatest imbalance will not be displayed as they usually are when a threshold is specified. 

Second, we see a balance summary across all the clusters. This table presents the minimum, mean, and maximum balance statistics for each variable across clusters. Setting `un = TRUE` will also display the same values for the adjusted data set. With binary treatments, setting `disp.v.ratio = TRUE` or `v.threshold` to some number will display the same values for variance ratios. Setting `abs = TRUE` requests summaries of absolute balance statistics which displays the extremeness of balance statistics for each variable; thus, if, for example, in some groups there are large negative mean differences and in other groups there are large positive mean differences, this table will display large mean differences, even though the average mean difference is close to 0. While it's important to know the average balance statistic overall, assessing the absolute balance statistics provides more information about balance within each cluster rather than in aggregate.

To examine balance for just a few clusters at a time, users can enter values for `which.cluster`. This can be a vector of clusters indices (i.e., 1, 2, 3, etc.) or names (e.g., "black", "hispan", "white"). Users can also specify `which.cluster = .none` to omit cluster balance for all clusters and just see the summary across clusters, which might be helpful if there are many clusters. Users can also specify whether they want to see the summary across clusters by specifying `TRUE` or `FALSE` for `cluster.summary`. When `which.cluster = .none`, `cluster.summary` will automatically be set to `TRUE` (or else there wouldn't be any output!). When examining balance within a few groups, it can be more helpful to examine balance within each group and ignore the summary. Below are examples of the use of `which.cluster` and `cluster.summary` to change `bal.tab()` output.

```{r}
#Just for black and hispan
bal.tab(m.out, cluster = "race", which.cluster = c("black", "hispan"),
        cluster.summary = FALSE)

#Just the balance summary across clusters with only the mean
bal.tab(m.out, cluster = "race", which.cluster = .none, cluster.fun = "mean")
```

These can also be set as global options by using, for example, `set.cobalt.options(cluster.fun = "mean")`, which allows user not to type a non-default option every time they call `bal.tab`.

### `bal.plot()`

`bal.plot()` functions as it does with non-clustered data, except that multiple plots can be produced at the same time displaying balance for each cluster. The arguments to `bal.plot()` are the same as those for `bal.tab()`, except that `cluster.summary` is absent. Below is an example of the use of `bal.plot()` with clustered data:

```{r}
bal.plot(m.out, var.name = "age", cluster = "race", which = "both")
```

Balance plots for each cluster are displayed next to each other. You can specify `which.cluster` as with `bal.tab()` to restrict plotting to a subset of clusters.

### `love.plot()`

`love.plot()` shines with clustered data because there are several options that are unique to `cobalt` and help with the visual display of balance. One way to display cluster balance with `love.plot()` is to produce different plots for each cluster, as `bal.plot()` does. This method should not be used with many clusters, or the plots will be unreadable. In our present example, this is not an issue. To do so, the `which.cluster` argument in `bal.tab()` or `love.plot()` must be set to the names or indices of the clusters for which balance is to be plotted. If `which.cluster` is set to `.all` (the default), all clusters will be plotted. Below is an example:

```{r}
love.plot(m.out, cluster = "race")
```

These plots function like those from using `love.plot()` with non-clustered data, except that they cannot be sorted based on the values of the balance statistics (they can still be sorted alphabetically, though). This is to ensure that the covariates line up across the plots. The same axis limits will apply to all plots.

Second, balance can be displayed summarizing across clusters by plotting an aggregate function (i.e., the mean or maximum) of the balance statistic for each covariate across clusters. To do this, `which.cluster` in the `love.plot` command must be set to `.none`. To change which aggregate function is displayed, use the argument to `agg.fun`, which may be "mean" or "max". Below is an example:

```{r}
love.plot(m.out, cluster = "race", which.cluster = .none, agg.fun = "mean")
```

A third option is to set `agg.fun = "range"` (the default), which produces a similar plot as above except that the minimum and maximum values of the balance statistics for each covariate are displayed as well. See below for an example:

```{r}
love.plot(m.out, cluster = "race", which.cluster = .none, agg.fun = "range")
```

Each point represents the mean balance statistic, and the bars represent intervals bounded by the minimum and maximum of each balance statistic. This display can be especially helpful with many clusters given that the mean alone may not tell the whole story. In some cases, it might be useful to set limits on the x-axis by using the `limits` argument in `love.plot()`; doing so may cut off some of the ranges, but whatever is left will be displayed. All `love.plot()` arguments work with these methods as they do in the case of non-clustered data. When `var.order` is specified as `"unadjusted"` or `"adjusted"`, the ordering will occur on the mean balance statistic when using `agg.fun = "range"`. Only one argument to `stats` is allowed when segmented data produces more than one plot (i.e., as it would with `which.cluster = .all`).

## Multiply Imputed Data

Multiply imputed data works in a very similar way to clustered data, except the "grouping" variable refers to imputations rather than clusters. Thus, each row belongs to one imputation (i.e., the data set should be in "long" format). The data set used should only include the imputed data sets and not the original data set with missing values (unlike Stata's `mi` commands, which do require the original data set as well). The imputed data sets can be of different sizes (i.e., because matching reduced the size of each differently), but it is preferred that they are the same size and weights are used to indicate which units belong to the sample and which do not.

In the example below, we will use a version of the Lalonde data set with some values missing. We will use the `mice` package to implement multiple imputation with chained equations. We will perform the "within" approach using the `MatchThem` to perform propensity score weighting within each imputation with `educ` as the continuous treatment (substantively this analysis makes no sense and is just for illustration).

```{r}
library("mice"); library("MatchThem"); library("cobalt")
data("lalonde_mis", package = "cobalt")

#Generate imputed data sets
m <- 10 #number of imputed data sets
imp.out <- mice(lalonde_mis, m = m, print = FALSE) 


#Performing generalized propensity score weighting in each imputation
wt.out <- weightthem(educ ~ age + race + married + 
                      re74 + re75, datasets = imp.out, 
                     approach = "within", method = "ps")
```

### `bal.tab()`

There are a few ways to use `bal.tab()` with our imputed data sets. When using the `mimids` or `wimids` methods for `MatchThem` objects, only the output object needs to be supplied. When using other methods, an argument to `imp` can be supplied; this should contain the imputation identifiers for each unit or the name of a variable in a supplied dataset (e.g., through the `data` argument) that contains the imputation identifiers. Alternatively, the `mids` object resulting from the call to `mice` can be supplied to the `data` argument, which automatically populates `imp`. There are four arguments that are only relevant to imputed data:

* `imp` is a vector of imputation numbers for each unit or the name of a variable in an available data set containing the imputation numbers. If `data` is a `mids` object or if the `mimids` or `wimids` methods are used, this doesn't need to be specified.

* `which.imp` determines for which imputation balance assessment is to be displayed. Often it can be useful to examine balance in just a few imputations for a detailed examination of what is going on. Can be `.all` to display all imputations (not recommended), `.none` to display none, or a vector providing the imputation numbers for the desired imputations. (Default: no imputations are displayed.)

* `imp.summary` determines whether to display a summary of balance across imputations. (Default: the summary of balance across imputations is displayed.)

* `imp.fun` determines which function(s) are used to combine balance statistics across imputations for the summary of balance across imputations. (Default: when `abs = FALSE`, minimum, mean, and maximum; when `abs = TRUE`, mean and maximum)

`which.imp`, `imp.summary`, and `imp.fun` affect display only and can also be set as global options by using `set.cobalt.options()` like the corresponding cluster options.

In many cases, not all variables are imputed, and often the treatment variable is not imputed. If each imputation has the same number of units, you can specify other arguments (e.g., treatment, distance) by specifying an object of the length of one imputation, and this vector will be applied to all imputations. This will come in handy when supplying additional covariates that weren't involved in the imputation or propensity score estimation through `addl`. To do this, the imputed data set must be sorted by imputation and unit ID.

Because we're using a `wimids` object, we can just call `bal.tab()` with it as the first argument.

```{r}
#Checking balance on the output object
bal.tab(wt.out)
```

First, we see a balance summary across all the imputations. This table presents the minimum, mean, and maximum balance statistics for each variable across imputations. Setting `un = TRUE` will also display the same values for the adjusted data set. Setting `abs = TRUE` will make `bal.tab` report summaries of the absolute values of the balance statistics. This table functions in the same way as the table for balance across clusters. Below is the average sample size across imputations; in some matching and weighting schemes, the sample size (or effective sample size) may differ across imputations.

To view balance on individual imputations, you can specify an imputation number to `which.imp`. To hide the balance summary across imputations, you can specify `imp.summary = FALSE`. 

```{r}
bal.tab(wt.out, which.imp = 1, imp.summary = FALSE)
```

As with clustered data, all `bal.tab()` options work as with non-imputed data. Indeed, the functions for clustered and imputed data are nearly identical except that for imputed data, `bal.tab()` computes the average sample size across imputations, whereas for other forms of segmented data, `bal.tab()` computes the total sample size across groups.

### `bal.plot()`

`bal.plot()` works with imputed data as it does with non-imputed data, except that multiple plots can be produced displaying balance for multiple imputations at a time. The arguments to `bal.plot()` are the same as those for `bal.tab()`, except that `imp.summary` is absent. Below is an example of the use of `bal.plot()` with imputed and matched data from `MatchThem`, examining balance in the first imputation:

```{r}
bal.plot(wt.out, which.imp = 1, var.name = "age", which = "both")
```

When many imputations are generated, it is recommended not to plot all at the same time by specifying an argument to `which.imp`, as done above.

### `love.plot()`

`love.plot()` functions with imputed data as it does with clustered data. It is not recommended to display balance for multiple imputations at a time, and rather to display balance summarized across imputations:

```{r}
love.plot(wt.out, threshold = .05)
```

Often these ranges will be small if the imputed data sets are very similar to each other, but the more imputations are generated, the wider the ranges tend to be.

## Multi-Category Treatments with Multiply Imputed Data

So far we've seen how `cobalt` functions work with one layer of data segmentation, but now let's see what it's like to work with two or more layers of segmentation. As an example, we'll first look at multiply imputed data with a multi-category treatment. With multi-category treatments, balance is typically assessed by examining balance statistics computed for pairs of treatments. With multi-category and multiply imputed data, the data is segmented both by imputation and by treatment pair.

We'll use the three-category `race` variable as our multi-category treatment and use the same imputed data as above. Again, the `MatchThem` package can be used to estimate weights in multiply imputed data. We'll use propensity score weighting to estimate the ATE of `race`. As before, this analysis makes no sense substantively and is just for illustration.

```{r}
#Estimate weights within each imputation using propensity scores
wt3.out <- weightthem(race ~ age + educ + married + 
                      nodegree + re74 + re75, 
                     datasets = imp.out, approach = "within", 
                     method = "ps", estimand = "ATE")
```

### `bal.tab()`

Using `bal.tab()` on the resulting object does the following: for each pair of treatments, balance is assessed for each imputation and aggregated across imputations. That is, for each pair of treatments, everything described in the previous section will occur. 

```{r}
bal.tab(wt3.out)
```

Other options can be supplied to choose how balance is computed with multi-category treatments; these are described at `?bal.tab.multi` and in the main vignette. Importantly, though, a balance summary across treatment pairs is not available, though this may change in future releases. 

### `bal.plot()`

`bal.plot()` works with multi-category treatments the same way it does with binary treatments. All treatment levels are displayed on the same plot. As before, with multiply imputed data, balance can be examined on one or more imputations at a time.

```{r}
bal.plot(wt3.out, var.name = "married", which.imp = 1,
         which = "both")
```

### `love.plot()`

With multiple layers of segmentation, `love.plot()` has a few options. Before, we saw that we could facet the plot by the segments or aggregate across segments; with multiple layers, we can do both. `love.plot()` can aggregate across as many layers as there are and can facet with segments of one layer. With more than two layers of segmentation, at least one of the `which.` arguments must be `.none` (to aggregate) or of length 1 (to facet at one segment of the layer). Here we'll demonstrate aggregating across imputations while faceting on treatment pairs.

```{r}
love.plot(wt3.out, threshold = .1, agg.fun = "mean")
```

The arguments to `which.treat`, `which.imp`, `abs`, and `agg.fun` can be used to control how the plots are faceted and aggregated as they can with single-layer data.

## Concluding Remarks

We have demonstrated the use of `cobalt` with clustered data, multiply imputed, and multiply imputed data with a multi-category treatment. Though there are few published recommendations for the display of balance in some of these cases, we believe these tools may encourage development in this area. In general, we believe in displaying the most relevant information as compactly as possible, and thus recommend using `love.plot()` with some degree of aggregation for inclusion in published work. 